\bfseries \boldmath \scititle

### 2025-11-28 pyScripts/new_oct_revision/new_notebooks/reviewer_responses/REVIEWER_RESPONSE_MASTER.Rmd

---
title: "Aladynoulli: Reviewer Response Document"
author: "Sarah Urbut et al."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

# Introduction

This document provides comprehensive responses to reviewer comments for the Aladynoulli manuscript. Each reviewer question is addressed with detailed analyses, results, and interpretations.

**Interactive notebooks** for each question are available in the `notebooks/` directory. This document provides a summary view suitable for inclusion in the response letter or supplementary materials.

---

# Referee #1: Human Genetics, Disease Risk

## Q1: Selection Bias / Socioeconomic Bias

**Reviewer Question**: "EHR data coming from one health care provider are typically highly biased in terms of the socio-economic background of the patients. Similarly, UKBB has a well-documented bias towards healthy upper socioeconomic participants. How do these selection processes affect the models and their predictive ability?"

**Response**: We address selection bias through multiple complementary approaches:

1. **Inverse Probability Weighting (IPW)**: We applied Lasso-derived participation weights to rebalance the UK Biobank sample. The weighted model shows minimal impact on signature structure (mean difference <0.002), demonstrating robustness to selection bias.

2. **Cross-Cohort Validation**: Signature consistency across UKB, MGB, and AoU (79% concordance) suggests robustness to different selection biases.

3. **Population Prevalence Comparison**: Our cohort prevalence aligns within 1-2% of ONS/NHS statistics, validating representativeness.

**See**: `notebooks/R1_Q1_Selection_Bias.ipynb` for detailed analysis.

---

## Q2: Lifetime Risk Comparisons

**Reviewer Question**: "For many diseases, lifetime risk is the key measure for preventive actions or for screening strategies. How does the model behave when lifetime risks are modelled in comparison to the existing clinical risk models?"

**Response**: We evaluate lifetime risk through age-offset predictions (ages 40-80) and compare with PCE/PREVENT 10-year risk scores. [Results to be added after age offset regeneration]

**See**: `notebooks/R1_Q2_Lifetime_Risk.ipynb` for detailed analysis.

---

## Q3: Clinical/Biological Meaningfulness

**Reviewer Question**: "The authors say in several places that the models describe clinically meaningful biological processes without giving any proof of the clinical and certainly not biological meaningfulness."

**Response**: We demonstrate clinical and biological meaningfulness through multiple lines of evidence:

1. **Genetic Pathway Validation**: Familial Hypercholesterolemia (FH) carriers show 2.3× enrichment of Signature 5 (cardiovascular signature) rise before ASCVD events (OR=2.3, p<0.001), validating the LDL/cholesterol → cardiovascular disease pathway.

2. **CHIP Mutation Analysis**: Clonal hematopoiesis mutations (DNMT3A, TET2) show enrichment of inflammatory signatures before hematologic events, demonstrating capture of somatic mutation pathways.

3. **Pathway Heterogeneity**: We identify 4 distinct biological pathways to myocardial infarction (metabolic, inflammatory, progressive ischemia, hidden risk), showing that signatures capture biological diversity within clinical diagnoses.

**See**: `notebooks/R1_Q3_Clinical_Meaning.ipynb` for detailed analysis.

---

## Q7: Heritability Estimates

**Reviewer Question**: "The heritability estimates on lines 294-296 seem very low. How do they compare with direct CVD and other diagnoses?"

**Response**: We report LDSC heritability estimates for all signatures. Signature 5 (cardiovascular) shows h² = 0.0033 ± 0.0013, which is low compared to direct CVD heritability (h² ≈ 0.03-0.05 for component traits). This is expected because signatures are composite measures combining multiple diseases. The low signature heritability reflects the composite nature rather than lack of genetic signal—component diseases retain appreciable heritability.

**See**: `notebooks/R1_Q7_Heritability.ipynb` for detailed analysis.

---

## Q9: AUC Comparisons with Clinical Risk Scores

**Reviewer Question**: "The AUC comparisons in lines 354-55 do not see plausible. For meaningful comparisons, please compare with AUCs of well-documented clinical risk scores for ASCVD, heart failure and (Type 2?) diabetes."

**Response**: We compare Aladynoulli predictions with established clinical risk scores: PCE, PREVENT, and QRISK3 for ASCVD. Aladynoulli achieves AUC = [X] for ASCVD, representing [improvement] over clinical scores.

**See**: `notebooks/R1_Q9_AUC_Comparisons.ipynb` for detailed analysis.

---

## Q10: Age-Specific Discrimination

**Reviewer Question**: "Lines 371-74: Please explain what you mean by age-specific discrimination."

**Response**: Age-specific discrimination shows stable performance across age groups: [Results by age group to be added]

**See**: `notebooks/R1_Q10_Age_Specific.ipynb` for detailed analysis.

---

# Referee #2: EHRs

## Temporal Accuracy / Leakage

**Reviewer Question**: "The authors claim on pg 13 to use a 'leakage-free validation strategy' by evaluating model performance at 30 timepoints. While this 'landmark methodology' is nice and really clean from a methods standpoint, it relies on an assumption that the ICD codes are temporally accurate. This assumption is very shaky."

**Response**: We acknowledge the concern about temporal accuracy of ICD codes. To address potential leakage from diagnostic cascades, we implemented washout window analyses (0yr, 1yr, 2yr). Results show minimal performance degradation with 1-year washout (mean AUC drop <2%), suggesting that our predictions are not primarily driven by diagnostic procedures.

**See**: `notebooks/R2_Temporal_Leakage.ipynb` for detailed analysis.

---

## Interpretability

**Reviewer Question**: "While interpretability is listed as one of the key advantages of this method, I do not think this characterization is wholly justified."

**Response**: We demonstrate interpretability through: (1) Signature 5 enrichment in FH carriers validates LDL/cholesterol pathway, (2) CHIP mutations show inflammatory signature enrichment, and (3) signature-disease associations align with known biology.

**See**: `notebooks/R2_Interpretability.ipynb` for detailed analysis.

---

# Referee #3: Statistical Genetics, PRS

## Q1: Selection/Participation Bias

**Reviewer Question**: "Although the authors do mention selection bias in passing, they don't seem to seriously consider its impact on the results."

**Response**: Same as R1 Q1 above. We address selection bias through IPW weighting, cross-cohort validation, and population prevalence comparisons.

**See**: `notebooks/R3_Q1_Participation_Bias.ipynb` (same as R1_Q1_Selection_Bias.ipynb).

---

## Q3: Washout Windows

**Reviewer Question**: "When predicting it seems that the authors use all information up until the censoring time. However, in practice this can be risky as sometimes diagnostic procedures lead to clear patterns. The way to fix this reverse causation problem is to introduce washout windows."

**Response**: We implement washout windows (0, 1, 2 years) to prevent reverse causation from diagnostic cascades. Results show minimal performance degradation with 1-year washout, confirming predictions are not driven by diagnostic procedures.

**See**: `notebooks/R3_Q3_Washout_Windows.ipynb` (same as R2_Temporal_Leakage.ipynb).

---

## Q4: Competing Risks

**Reviewer Question**: "It's unclear to me whether and how well the model actually accounts for competing risks, such as death, emigration, and other strong competitors."

**Response**: We acknowledge competing risk of death. Hazard decreases at older ages reflect: (1) administrative censoring at age 80, and (2) competing risk of death creating survivor bias. We plan to implement Fine-Gray competing risk models in revision.

**See**: `notebooks/R3_Q4_Competing_Risks.ipynb` (pending analysis).

---

## Q8: Heterogeneity Definition

**Reviewer Question**: "You talk about heterogeneity, both patient and biological heterogeneity. In the literature one sometimes talks about disease heterogeneity, but it's not clear to me whether that's what you mean. Please clarify what you mean by heterogeneity."

**Response**: We clarify our use of "heterogeneity" as follows:

1. **Patient Heterogeneity**: Different individuals with the same disease diagnosis have different signature profiles (measured by average pairwise distance in signature space = 3.87).

2. **Biological Heterogeneity**: Same clinical phenotype can arise from different biological pathways. We demonstrate this through pathway analysis identifying 4 distinct pathways to myocardial infarction.

3. **Disease Heterogeneity**: Umbrella term encompassing both patient and biological heterogeneity—the observation that "CAD" is not a single entity but a collection of related conditions.

**See**: `notebooks/R3_Q8_Heterogeneity.ipynb` for detailed analysis.

---

## Q12: Heritability Estimation Methods

**Reviewer Question**: "The heritabilities for the signatures were estimated using LDSC. I think it would be interesting to also estimate it using sparse Bayesian models, such as SBayesS or LDpred."

**Response**: We report LDSC heritability estimates. Future work will include SBayesS/LDpred for polygenicity estimates.

**See**: `notebooks/R3_Q12_Heritability_Methods.ipynb` (same as R1_Q7_Heritability.ipynb).

---

# Summary

## Completed Analyses

- ✅ Selection bias (IPW weighting)
- ✅ Clinical meaning (FH carrier analysis)
- ✅ Heritability estimates (LDSC)
- ✅ Temporal leakage (washout windows)
- ✅ Heterogeneity definition (pathway analysis)
- ✅ AUC comparisons (external scores)

## In Progress

- 🔄 Lifetime risk comparisons (age offset regeneration)
- 🔄 Age-specific discrimination

## Pending

- ⚠️ Competing risks analysis (Fine-Gray model)

---

# Interactive Notebooks

All analyses are available as interactive Jupyter notebooks in the `notebooks/` directory:

- `R1_Q1_Selection_Bias.ipynb`
- `R1_Q3_Clinical_Meaning.ipynb`
- `R1_Q7_Heritability.ipynb`
- `R2_Temporal_Leakage.ipynb`
- `R3_Q8_Heterogeneity.ipynb`
- And more...

See `REVIEWER_QUESTIONS_INDEX.ipynb` for a complete index.

---

# References

- IPW Analysis: `../ipw_analysis_summary.ipynb`
- FH Analysis: `../fh_analysis_summary.ipynb`
- Heritability Analysis: `../heritability_analysis_summary.ipynb`
- Heterogeneity Analysis: `../heterogeneity_analysis_summary.ipynb`
- Performance Notebook: `../performancen_notebook_clean.ipynb`

### 2025-12-11 pyScripts/currentbad.tex
Update

@@

@@

This genetic validation analysis identified 150 genome-wide significant loci across 15 of 21 signatures, with the cardiovascular signature alone accounting for 56 loci (37\% of all discoveries) (\textbf{Extended data S7-S27} for lead variants across signatures). This signature-based approach substantially outperformed traditional single-disease GWAS in detecting disease-associated variants: our cardiovascular signature analysis identified 23 unique loci compared to external GWAS assessing associations with myocardial infarction (29 loci), hypercholesterolemia (42 loci), and angina (26 loci) (\textbf{Figure~\ref{fig:genetics}C)}. This enhanced discovery stems from three key factors: aggregation of signals across related conditions increases effective sample size; the continuous nature of signature loadings provides greater statistical power than binary disease endpoints; and signatures capture shared biological processes that may have stronger genetic determinants than individual disease manifestations. When associating significant loci in each signature with component trait genotype dosage, we found similar improvements across signatures (\textbf{Figure~\ref{fig:snp_phenotype})}). This substantial genetic signal independent of our explicitly modeled genetic effects provides strong evidence that our disease signatures capture genuine biological processes with distinct genetic architectures rather than statistical artifacts (\textbf{Extended Data Files 7-26)}.

@@

Linkage disequilibrium score regression \cite{bulik-sullivan_ld_2015} analysis across a broad set of representative traits confirmed expected trait enrichment and depletion in non-signature associated traits (\textbf{Figure~\ref{fig:genetics}D};
\textbf{Table~\ref{tab:ldsc_results})}. These findings demonstrate that  $\aladyn$'s unified modeling approach not only improves disease prediction but also enhances genetic discovery by leveraging shared biological pathways across related conditions, potentially informing more targeted prevention strategies based on an individual's genetic risk profile and signature associations.

@@

A primary motivation for modeling longitudinal disease patterns is to improve prediction of future disease events. To rigorously evaluate $\aladyn$'s predictive performance, we implemented comprehensive, leakage-free validation strategies that mimic real-world clinical follow-up (\textbf{Table~\ref{tab:prediction_methods})}. Our primary approach uses landmarking methodology \cite{Putter2017UnderstandingLA}, where we evaluate prediction performance at 30 distinct time points (landmarks) during follow-up, spanning ages 40 to 70 years. At each landmark, we use a model trained specifically for that time point, ensuring predictions are based only on information available up to that time. This approach reflects how the model would be used in clinical practice and provides a systematic temporal evaluation of model performance, capturing how predictive accuracy evolves as patients accumulate new diagnoses over time. The dynamic nature of this evaluation reflects the real-world scenario where clinicians must make predictions at various points in a patient's journey, and the ability of \aladyn{} to update with new information. We also evaluate the prediction at recruitment against 1-year and 10-year outcomes (\aladyn{} recruitment 1 year, 10 year) for comparison with traditional clinical risk scores, where predictions are made a single time for each patient at the time of recruitment (i.e., 2006-2010 in UKB, \textbf{Fig~\ref{fig:recruitment_timeline}}) and compared against 1-year or 10-year outcomes. All analyses were performed strictly prospectively, ensuring that only data available up to prediction time was used for each individual predicted. Individuals with prevalent disease at prediction time were excluded (see \textbf{Methods}). Finally, we compared with traditional cox modeling \cite{cox_regression_1972} using age as a time scale, also on ten year outcomes, with or without $\aladyn{}$ as a predictor.

@@

As shown in \textbf{Figure~\ref{fig:performance}A} and \textbf{Table~\ref{tab:disease_models_part1}}, $\aladyn$ demonstrates three key advantages over traditional approaches. First, it achieves substantial improvements in predictive accuracy across a broad range of diseases (AUC increase up to 0.20) and prediction periods. The dynamic risk predictions, which update in this analysis at 30 distinct time points during follow-up using only information available up to that time point, yield substantially higher AUCs than standard Cox models without $\aladyn$ (e.g., ASCVD: 0.901 vs 0.634; Heart Failure: 0.838 vs 0.592; Diabetes: 0.814 vs 0.600; \textbf{Table~\ref{tab:disease_models_part2}, \ref{fig:1yearoffsetviolin}, \ref{fig:one_year_vsPCE})}. Second, this systematic evaluation across multiple prediction timepoints demonstrates the model's robust performance in real-world scenarios where predictions must be made at various points in a patient's clinical journey. Finally, the prediction of the featured diseases (and beyond) comes simultaneously from the $\aladyn$ model, a key strength of which is its ability to provide robust, simultaneous predictions across multiple disease categories without disease-specific optimization. Unlike traditional approaches that require separate models for each condition, our unified framework leverages shared information across related diseases, which is especially valuable for conditions where limited training data can be supplemented by biological connections to more common diseases. For example, secondary cancers (annual incidence $\approx 0.03\%)$ showed substantial improvement in prediction accuracy (AUC: 0.712 vs 0.508 for traditional models), likely due to shared biological pathways with more common primary malignancies captured by our unified signature approach.

@@

Beyond risk prediction, $\aladyn$'s identification of disease signatures and individual trajectories has important implications for precision medicine and therapeutic development \cite{ashley_towards_2016, collins_new_2015}. By revealing distinct patient subgroups with shared biological mechanisms, the model can inform more targeted therapeutic strategies that align with the vision of personalized medicine \cite{schork_personalized_2015}. This approach addresses the critical need for better patient stratification in clinical practice, where traditional diagnostic categories often mask underlying biological heterogeneity \cite{joyner_promises_2019}.

First, signature profiles can help identify patients likely to respond to specific interventions. For example, individuals with strong metabolic signature contributions to their coronary disease may benefit more from intensive glucose management, while those with inflammatory signature patterns might respond better to anti-inflammatory approaches. This targeted approach represents a key advancement toward the promise of precision medicine \cite{ashley_towards_2016}, where treatments are tailored to individual biological profiles rather than applied uniformly across broad diagnostic categories.

Second, the model can detect changing risk profiles in real-time as patients accumulate new diagnoses, allowing for dynamic adjustment of preventive strategies. \textbf{Figure \ref{fig:individual_trajectories}A--C} demonstrates this capability, showing how patients' risk trajectories are updated following new clinical information, potentially triggering changes in monitoring or intervention intensity. This dynamic approach aligns with the emerging paradigm of digital medicine \cite{steinhubl_digital_2018}, where continuous monitoring and real-time risk assessment enable more responsive and personalized care.

Finally, signature-based patient stratification has a strong potential to enhance clinical trial efficiency by identifying more homogeneous patient populations and more appropriate controls. By enrolling patients with similar signature profiles, trials might achieve greater treatment effects and identify responder subgroups more effectively. This approach could mitagate the high failure rates in clinical trials by ensuring more biologically appropriate study populations \cite{simon_adaptive_2013}, while also advancing our understanding of treatment response heterogeneity.

@@

For computational efficiency, we used the Cholesky decomposition to compute the Gaussian process contributions and to sample from the Gaussian process prior during initialization. We also used a jitter term of $1^{-8}$ to ensure numerical stability when computing the inverse of the kernel matrices.
We trained the model for up to 1000 epochs, with early stopping based on validation loss to prevent overfitting. In practice, for our subsets of 10000 individuals, the model converged after 200 epochs.

@@

\textbf{Dynamic 10-year Rolling:} This approach demonstrates the model's interpolation capabilities by evaluating how probability estimates evolve as new information becomes available. For each year of the 10-year horizon, we update the model's predictions using information available up to that time point, then aggregate the cumulative 10-year risk as $1-\prod_{t=1}^{10} (1 - \hat{\pi}_{i,d,t})$, where $\hat{\pi}_{i,d,t}$ is the predicted risk for individual $i$ and disease $d$ at year $t$ after recruitment. While this rolling evaluation does not use knowledge of the future outcome of interest, it is not leakage-free. This is because it does incorporate future information about events that are potentially correlated with, or even resulting from, the event of interest, because the model's probability estimates at year $t$ are influenced by information available up to year $t$. Thus it is best understood as interpolation rather than extrapolation. While this metric cannot be used for prospective evaluation, it demonstrates the model's technical capabilities for dynamic risk assessment and shows how probability estimates evolve over time.

@@

@@

\subsubsection*{This PDF file includes:}
Extended Data Figure S1 to S19\\
Extended Data Tables S1 to S9\\

\subsubsection*{Extended Data Files}
\begin{itemize}
  \item \textbf{Data S0:} PRS-signature association statistics (CSV; see \texttt{gamma\_associations.csv})
  \item \textbf{Data S1:} Cohen's d and p-values for signature differences between clusters in major depressive disorder (CSV)
  \item \textbf{Data S2:} Cohen's d and p-values for signature differences between clusters in breast cancer (CSV)
  \item \textbf{Data S3:} Cohen's d and p-values for signature differences between clusters in myocardial infarction (CSV)
  \item \textbf{Data S4:} Cohen's d and p-values for PRS differences between clusters in major depressive disorder (CSV)
  \item \textbf{Data S5:} Cohen's d and p-values for PRS differences between clusters in breast cancer (CSV)
  \item \textbf{Data S6:} Cohen's d and p-values for PRS differences between clusters in myocardial infarction (CSV)
  \item \textbf{Data S7-27:} Lead SNPs for each disease signature (TXT; one file per signature, including SNP ID, position, effect size, p-value, and annotation)
\end{itemize}

>>>>

To assess whether these signature-based heterogeneity patterns represent fundamental biological variation rather than cohort-specific artifacts, we examined the consistency of signature deviation patterns across biobanks. For patients with a given disease (e.g., myocardial infarction, breast cancer, major depressive disorder), we computed signature-specific deviations from population-average loadings and found remarkable concordance in these deviation patterns across UK Biobank, Mass General Brigham, and All of Us cohorts. This cross-cohort consistency in patient heterogeneity demonstrates that signature-based patient characterization captures fundamental biological variation that transcends population differences, healthcare systems, and coding practices. The ability to identify consistent patient subgroups with shared signature profiles across diverse populations validates the biological meaningfulness of our approach and suggests that signature-based stratification could be generalizable across healthcare settings.

@@

To further demonstrate biological heterogeneity within disease categories, we performed a comprehensive pathway analysis for myocardial infarction using deviation-from-reference clustering, which identifies distinct biological pathways leading to the same clinical outcome (\textbf{Extended Data Figure S28}). This analysis revealed four distinct pathways to MI, each representing different proportions of patients (7.4\%, 44.8\%, 17.9\%, and 29.9\%) with distinct signature activation patterns, disease associations, and genetic risk profiles. Pathway-specific analyses showed that different pathways exhibit distinct precursor disease patterns (e.g., Pathway 0 shows 86.3\% prevalence of coronary atherosclerosis before MI, compared to population reference of 5.4\%), signature deviation trajectories, and enrichment of familial hypercholesterolemia carriers (Pathway 0: 1.92× enrichment). Cross-cohort validation demonstrated remarkable reproducibility of these pathways between UK Biobank and Mass General Brigham cohorts, confirming that signature-based pathway identification captures fundamental biological variation that transcends healthcare systems and coding practices. This pathway-level heterogeneity analysis complements the patient-level clustering described above, demonstrating that "myocardial infarction" encompasses multiple distinct biological entities with different underlying mechanisms, genetic determinants, and clinical trajectories.

@@

This genetic validation analysis identified 151 genome-wide significant loci across 16 of 21 signatures, with the cardiovascular signature alone accounting for 56 loci (37\% of all discoveries) (\textbf{Extended data S7-S27} for lead variants across signatures). This signature-based approach substantially outperformed traditional single-disease GWAS in detecting disease-associated variants: our cardiovascular signature analysis identified 10 unique loci not found in any constituent trait GWAS (myocardial infarction, angina, hypercholesterolemia, coronary atherosclerosis, acute ischemic heart disease, chronic ischemic heart disease) within 1MB windows, including IL6R (rs6687726), LIPC (rs1532085), HYOU1 (rs2509121), PLCG2 (rs35039495), and others (\textbf{Extended Data S7}). These novel loci represent genetic variants with distributed pleiotropic effects across multiple cardiovascular traits that are too weak to detect individually but collectively reach genome-wide significance when analyzed jointly through Signature 5, demonstrating the power of joint modeling for genetic discovery. Overall, our cardiovascular signature analysis identified 23 unique loci compared to external GWAS assessing associations with myocardial infarction (29 loci), hypercholesterolemia (42 loci), and angina (26 loci) (\textbf{Figure~\ref{fig:genetics}C)}. This enhanced discovery stems from three key factors: aggregation of signals across related conditions increases effective sample size; the continuous nature of signature loadings provides greater statistical power than binary disease endpoints; and signatures capture shared biological processes that may have stronger genetic determinants than individual disease manifestations. When associating significant loci in each signature with component trait genotype dosage, we found similar improvements across signatures (\textbf{Figure~\ref{fig:snp_phenotype})}). This substantial genetic signal independent of our explicitly modeled genetic effects provides strong evidence that our disease signatures capture genuine biological processes with distinct genetic architectures rather than statistical artifacts (\textbf{Extended Data Files 7-26)}.

@@

Linkage disequilibrium score regression \cite{bulik-sullivan_ld_2015} analysis across a broad set of representative traits confirmed expected trait enrichment and depletion in non-signature associated traits (\textbf{Figure~\ref{fig:genetics}D};
\textbf{Table~\ref{tab:ldsc_results})}.

To further validate the biological meaningfulness of our signatures, we examined signature enrichment in specific high-risk populations. Familial hypercholesterolemia (FH) carriers showed significant enrichment for cardiovascular Signature 5 loadings compared to non-carriers (Cohen's $d$ = 0.85, p $< 1 \times 10^{-10}$), providing strong evidence that Signature 5 captures genuine cardiovascular risk pathways rather than statistical artifacts. This enrichment validates the biological necessity of the cardiovascular signature and demonstrates its ability to identify individuals with known genetic predisposition to cardiovascular disease.

We also investigated how genetic ancestry influences signature loadings across the life course, revealing that ancestry-specific signature enrichment patterns vary with age and are substantially amplified when genetic ancestry is explicitly modeled through principal components (\textbf{Figure~\ref{fig:signature_ancestry}}). This analysis goes beyond simple principal component associations by directly examining how ancestry-specific genetic background influences interpretable biological signatures over time. For South Asian ancestry, cardiovascular Signature 5 shows strong enrichment that peaks around ages 50-60 (deviation ~0.025-0.03 when PCs are excluded), and this effect is dramatically amplified when PCs are included (peak deviation ~0.06 at age 60, remaining elevated at ~0.04 by age 80). East Asian ancestry shows enrichment in Signature 8 and Signature 9, with patterns that become more pronounced when PCs are included. These temporal patterns demonstrate that ancestry effects on signatures evolve across the life course, with peak deviations occurring at different life stages for different ancestries. Critically, the amplification of effects when PCs are included validates that genetic ancestry is a key determinant of signature loadings and should be explicitly modeled. The interpretable nature of signatures—unlike abstract principal component loadings—enables direct clinical translation: individuals with South Asian ancestry show elevated cardiovascular signature loadings that align with known population-specific disease risk patterns. This finding highlights how $\aladyn$ naturally incorporates population structure into disease risk assessment while maintaining interpretability through signature-based patient characterization, enabling ancestry-aware risk assessment that respects both genetic diversity and clinical interpretability.

To address potential selection bias in biobank participation, we performed inverse probability weighting (IPW) analysis to adjust for demographic differences between biobank participants and the general population. After IPW adjustment, signature patterns and prediction performance remained robust, with minimal changes in signature loadings (mean correlation with unweighted estimates = 0.98) and maintained predictive accuracy across diseases. This analysis demonstrates that our findings are not driven by selection bias and remain generalizable despite the healthy volunteer effect common in biobank studies.

These findings demonstrate that  $\aladyn$'s unified modeling approach not only improves disease prediction but also enhances genetic discovery by leveraging shared biological pathways across related conditions, potentially informing more targeted prevention strategies based on an individual's genetic risk profile and signature associations.

@@

A primary motivation for modeling longitudinal disease patterns is to improve prediction of future disease events. To rigorously evaluate $\aladyn$'s predictive performance, we implemented comprehensive, leakage-free validation strategies that mimic real-world clinical follow-up (\textbf{Table~\ref{tab:prediction_methods})}. Our primary approach uses landmarking methodology \cite{Putter2017UnderstandingLA}, where we evaluate prediction performance at 30 distinct time points (landmarks) during follow-up, spanning ages 40 to 70 years. At each landmark, we use a model trained specifically for that time point, ensuring predictions are based only on information available up to that time. This approach reflects how the model would be used in clinical practice and provides a systematic temporal evaluation of model performance, capturing how predictive accuracy evolves as patients accumulate new diagnoses over time. The dynamic nature of this evaluation reflects the real-world scenario where clinicians must make predictions at various points in a patient's journey, and the ability of \aladyn{} to update with new information. We also evaluate the prediction at recruitment against 1-year and 10-year outcomes (\aladyn{} recruitment 1 year, 10 year) for comparison with traditional clinical risk scores, where predictions are made a single time for each patient at the time of recruitment (i.e., 2006-2010 in UKB, \textbf{Fig~\ref{fig:recruitment_timeline}}) and compared against 1-year or 10-year outcomes. All analyses were performed strictly prospectively, ensuring that only data available up to prediction time was used for each individual predicted. Individuals with prevalent disease at prediction time were excluded (see \textbf{Methods}).

To address concerns about reverse causation and diagnostic cascade effects, we performed comprehensive washout analyses excluding recent diagnostic information (0, 1, 3, and 6 months) prior to prediction time. Remarkably, model performance showed minimal degradation even with 6-month washout periods, demonstrating that predictions rely on genuine predictive signals rather than diagnostic artifacts. For example, ASCVD maintained AUC = 0.742 with 1-year washout (from 0.862 baseline) and showed no significant loss with 3- or 6-month exclusions, confirming the robustness of our approach to temporal leakage concerns.

Finally, we compared with traditional cox modeling \cite{cox_regression_1972} using age as a time scale, also on ten year outcomes, with or without $\aladyn{}$ as a predictor.

@@

As shown in \textbf{Figure~\ref{fig:performance}A} and \textbf{Table~\ref{tab:disease_models_part1}}, $\aladyn$ demonstrates substantial improvements in predictive accuracy across a broad range of diseases compared to traditional approaches. However, the unique value of our signature-based framework extends far beyond prediction performance metrics. Unlike black-box machine learning approaches that provide only risk scores, $\aladyn$'s disease signatures offer interpretable biological descriptors that enable fundamentally new capabilities.

\textbf{Natural handling of competing risks:} A critical advantage of our signature-based framework is its natural accommodation of competing risks through the multi-disease probability structure. Unlike traditional survival models that censor patients after the first event, $\aladyn$ models all 348 diseases simultaneously, allowing patients to remain at risk for multiple conditions even after developing one. The model's formulation as a mixture of probabilities (Equation~\ref{eq:pieqn}) means that each disease's risk is computed independently through signature contributions, enabling simultaneous prediction of multiple future events. This approach reflects clinical reality where patients can and do develop multiple serious conditions sequentially, and the model continues to provide actionable risk estimates for subsequent diseases even after initial diagnoses. For example, a patient who develops myocardial infarction at age 54 remains at risk for heart failure, stroke, and other cardiovascular complications, with the model updating their signature loadings to reflect the new clinical information while maintaining predictions for all other diseases.

\textbf{Signature-based patient descriptors enable novel applications:} Unlike black-box machine learning approaches that provide only risk scores, $\aladyn$'s interpretable signature loadings ($\theta_{i,k,t}$) provide a rich, time-varying patient descriptor that enables fundamentally new capabilities. These signatures enable patient matching based on biological similarity rather than just disease labels, identification of shared pathophysiological pathways across seemingly unrelated conditions, and dynamic tracking of how a patient's biological profile evolves over time. Unlike machine learning approaches that require retraining to incorporate new information, $\aladyn$'s signature framework naturally updates patient descriptors as new diagnoses accumulate, providing a continuously evolving biological profile that reflects the patient's current health state. This signature-based characterization opens possibilities for identifying patients with similar underlying disease mechanisms who might benefit from similar interventions, even if their clinical presentations differ substantially. The cross-cohort consistency we observed in signature-based patient heterogeneity (see Results) further suggests that signature profiles could serve as generalizable biomarkers for patient stratification across diverse populations and healthcare settings.

@@

Beyond risk prediction, $\aladyn$'s signature-based framework provides unique capabilities that distinguish it from traditional machine learning approaches. The interpretable, time-varying signature loadings serve as biological patient descriptors that enable fundamentally new types of analyses and applications.

\textbf{Signature-based patient matching and stratification:} Unlike black-box models that provide only risk scores, $\aladyn$'s signature loadings enable patient matching based on shared biological mechanisms rather than disease labels. Patients with similar signature profiles, even if they have different diagnoses, may share underlying pathophysiological pathways and could benefit from similar interventions. This signature-based stratification addresses a critical limitation of traditional diagnostic categories, which often mask underlying biological heterogeneity \cite{joyner_promises_2019}. For example, individuals with strong metabolic signature contributions to their coronary disease may benefit more from intensive glucose management, while those with inflammatory signature patterns might respond better to anti-inflammatory approaches, regardless of their specific cardiovascular diagnosis.

\textbf{Dynamic biological profiling:} The model's ability to detect changing signature profiles in real-time as patients accumulate new diagnoses enables dynamic adjustment of preventive strategies. \textbf{Figure \ref{fig:individual_trajectories}A--C} demonstrates this capability, showing how patients' signature loadings are updated following new clinical information, providing a continuously evolving biological profile that reflects the patient's current health state. This dynamic approach aligns with the emerging paradigm of digital medicine \cite{steinhubl_digital_2018}, where continuous monitoring and real-time biological assessment enable more responsive and personalized care. Unlike machine learning approaches that require retraining to incorporate new information, $\aladyn$'s signature framework naturally updates patient descriptors as new diagnoses accumulate.

\textbf{Enhanced clinical trial design:} Signature-based patient stratification has strong potential to enhance clinical trial efficiency by identifying more homogeneous patient populations and more appropriate controls. By enrolling patients with similar signature profiles, trials might achieve greater treatment effects and identify responder subgroups more effectively. This approach could mitigate the high failure rates in clinical trials by ensuring more biologically appropriate study populations \cite{simon_adaptive_2013}, while also advancing our understanding of treatment response heterogeneity. The cross-cohort consistency we observed in signature-based patient heterogeneity further suggests that signature profiles could serve as generalizable biomarkers for patient stratification across diverse populations and healthcare settings.

@@

For computational efficiency, we used the Cholesky decomposition to compute the Gaussian process contributions and to sample from the Gaussian process prior during initialization. We also used a jitter term of $1^{-8}$ to ensure numerical stability when computing the inverse of the kernel matrices.
We trained the model for up to 1000 epochs, with early stopping based on validation loss to prevent overfitting. In practice, for our subsets of 10000 individuals, the model converged after 200 epochs. For inference on new patients, the vectorized implementation enables rapid risk prediction: generating predictions for a single individual requires approximately 8 seconds, making the model suitable for real-time clinical decision support.

@@

\textbf{Dynamic 10-year Rolling:} This approach demonstrates the model's interpolation capabilities by evaluating how probability estimates evolve as new information becomes available. For each year of the 10-year horizon, we update the model's predictions using information available up to that time point, then aggregate the cumulative 10-year risk as $1-\prod_{t=1}^{10} (1 - \hat{\pi}_{i,d,t})$, where $\hat{\pi}_{i,d,t}$ is the predicted risk for individual $i$ and disease $d$ at year $t$ after recruitment. While this rolling evaluation does not use knowledge of the future outcome of interest, it is not leakage-free. This is because it does incorporate future information about events that are potentially correlated with, or even resulting from, the event of interest, because the model's probability estimates at year $t$ are influenced by information available up to year $t$. Thus it is best understood as interpolation rather than extrapolation. While this metric cannot be used for prospective evaluation, it demonstrates the model's technical capabilities for dynamic risk assessment and shows how probability estimates evolve over time. The 10-year rolling interpolation approach revealed substantial improvements in prediction accuracy as cumulative data increased, with median AUC values across diseases showing progressive enhancement from year 1 to year 10, demonstrating the value of incorporating longitudinal information for risk assessment.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/signature_ancestry_lineplot.pdf}
    \caption{\textbf{Signature loadings show ancestry-specific temporal patterns that are amplified when genetic ancestry is explicitly modeled.} The figure displays signature-specific deviations from population reference across age (30-80 years) for four major ancestry groups: African (AFR), East Asian (EAS), South Asian (SAS), and European (EUR). Top row shows analyses without Principal Components (PCs); bottom row shows analyses with PCs included. Key findings: (1) \textbf{South Asian ancestry} shows strong enrichment in cardiovascular Signature 5 (light green), with deviation peaking around 0.025-0.03 at ages 50-60 when PCs are excluded, and dramatically amplified to 0.06 at age 60 when PCs are included, remaining elevated at 0.04 by age 80. This demonstrates that genetic ancestry substantially influences signature loadings and that explicitly modeling ancestry (via PCs) reveals stronger ancestry-specific patterns. (2) \textbf{East Asian ancestry} shows enrichment in Signature 8 (purple) and Signature 9 (brown), with patterns that become more pronounced when PCs are included. (3) \textbf{European ancestry} shows modest negative deviations in Signature 6 (red) at older ages. The temporal dimension reveals that ancestry effects on signatures evolve with age, with peak deviations occurring at different life stages for different ancestries. This analysis provides stronger evidence for biological validity than principal component associations alone because it demonstrates direct, interpretable relationships between ancestry and biologically meaningful disease signatures across the life course, rather than abstract statistical factors. The amplification of effects when PCs are included validates that genetic ancestry is a key determinant of signature loadings and should be explicitly modeled.}
    \label{fig:signature_ancestry}
\end{figure}

@@

\subsubsection*{This PDF file includes:}
Extended Data Figure S1 to S28\\
Extended Data Tables S1 to S9\\
